---
title: "Title"
author: "Author"
date: 
output:
  html_document:
    theme: simplex
    toc: true
    toc_depth: 3
    keep_md: false
    toc_float: true
    code_folding: hide
---

<!-- These two chunks should be added in the beginning of every .Rmd that you want to source an .R script -->
<!--  The 1st mandatory chunck  -->
<!--  Set the working directory to the repository's base directory -->
```{r, echo=F, message=F} 
#Don't combine this call with any other chunk 
# cat("Working directory: ", getwd()) # check where you are
  library(knitr)
# Rmd is in "./reports/reportA/reportA.Rmd", but now the root is "./"
  knitr::opts_knit$set(root.dir='../../../') 
```

<!--  The 2nd mandatory chunck  -->
<!-- Set the report-wide options, and point to the external code file. -->
```{r set_options, echo=F}
# set options shared by all chunks
opts_chunk$set(
  results='show', 
  message = TRUE,
  comment = NA, 
  tidy = FALSE,
  fig.height = 4, 
  fig.width = 5.5, 
  out.width = "550px",
  fig.path = 'figure_rmd/',     
  dev = "png",
  dpi = 400
  # fig.path = 'figure_pdf/',     
  # dev = "pdf"#,
  # dev.args=list(pdf = list(colormodel = 'cmyk'))
)
echoChunks <- TRUE
options(width=120) #So the output is 50% wider than the default.
# connect to the file with the underlying R script  
read_chunk("./analysis/blogposts/basic-exploration/basic-exploration.R") 
```


# Abstract

In Progress 

<!-- Load packages to be used in this report. --> 
```{r load-packages, echo=echoChunks, message=FALSE}
```


<!-- Load the sources.  Suppress the output when loading sources. --> 
```{r load-sources, echo=echoChunks, message=FALSE}
```


<!-- Load any Global functions and variables declared in the R file.  Suppress the output. --> 
```{r declare-globals, echo=echoChunks, results='show', message=FALSE}
```

<!-- Declare any global functions specific to a Rmd output.  Suppress the output. --> 
```{r, echo=echoChunks, message=FALSE}
#Put code in here.  It doesn't call a chunk in the codebehind file.
```

# Data

The data for this exploration comes from several different sources:

1. The Diabetes data was taken from the US Diabetes Surveillance System; www.cdc.gov/diabetes/data; Division of Diabetes Translation - Centers for Disease Control and Prevention.  The data was downloaded by year, and compiled into one set.

2. The list of rural counties is compiled from The US Census Bureau, it includes all North Carolina Counties that are at least 90% rural, more on the topic can be found here [Rural America](https://gis-portal.data.census.gov/arcgis/apps/MapSeries/index.html?appid=7a41374f6b03456e9d138cb014711e01)

<!-- Load the datasets.   -->
```{r load-data, echo=echoChunks, results='show', message=FALSE, warning=FALSE}
```



# Data Tweaks

The Diabetes data comes quite tidy from the CDC, the script to combine each year can be found here <!-- INSERT LINK TO GIT HUB SCRIPT -->
The only tweaks done here are to combine the rural countines column, and the map data for creating maps.

<!-- Tweak the datasets.   -->
```{r tweak-data, echo=echoChunks, results='show', message=FALSE}
```


# Overall

When examining North Carolina as a whole we can see that NC has been trending much higher than the United States as a whole.  We see that in 2016 there was a large spike in diagnosied cases, unfortunally this is the last year of data available to see if this upward trend contiunes. 

```{r g1, echo=echoChunks, results='show', message=FALSE, fig.width=9, out.width="900px"}
```





session information
===========================================================================

For the sake of documentation and reproducibility, the current report was rendered in the following environment.  Click the line below to expand.

<details>
  <summary>Environment <span class="glyphicon glyphicon-plus-sign"></span></summary>
```{r session-info, echo=FALSE}
if( requireNamespace("devtools", quietly = TRUE) ) {
  devtools::session_info()
} else {
  sessionInfo()
} 
```
